MCML with Boundary Surfaces

### 2019-06-24 docs/report/article.tex
Finish report

\documentclass[]{article}
\usepackage[english]{babel} % advanced multi-language features
\usepackage{graphicx} % figures
\usepackage{hyperref} % references and URLs are clickable
\usepackage{listings} % simple code listings (monospaced font, line oriented, no syntax highlighting)
\usepackage[section]{placeins} % restricting floating elements like figures with FloatBarriers
\bibliographystyle{IEEEtranS}

@@

\begin{abstract}
	This project report describes an extended implementation of multi-layered Monte Carlo photon transport\cite{wang1992monte} (MCML). It supports structured boundaries, whereas original MCML only supported planar layers. This allows to use the method for simulation of light in more realistic materials. The report presents and compares radiance profiles of different boundary shapes. The focus lies on the properties reflectance and transmittance, that are important for visual appearance.
\end{abstract}

@@

The purpose of the MCML implementation presented here, is to provide a ground truth of surface appearance for physically based rendering. Monte Carlo methods are accepted to be leading in accuracy, but low-noise results have to be bought with long computation times. The implementation produces radiance profiles, that reveal the effect of varying surface shapes, and therefore allow to model surface appearance with approximate analytic curves. The curve equations, when found, can be used in faster rendering algorithms.

@@

Monte Carlo simulation of light is very accurate, because it simulates single photon packets and records the distribution of a large number of photons. This implies, that it is a statistical method, using a pseudo random number generator to compute, what can be thought of random walks of photons, visualized in \autoref{trajectory}.

@@

The material geometry in MCML is a sequence of layers as shown in \autoref{layer-model}. Photons enter through the coordinate origin and move in 3 dimensions.

@@

The work is based on the light propagation model by Prahl et al.~\cite{prahl89}. You can get an overview of the algorithm by looking at the flowchart in \autoref{flowchart}.

@@

Specifics with regard to multiple layers, variable boundary shapes and general implementation are described in \autoref{impl}.

@@

The other contribution is support for non-planar layers. This is done by representing boundaries as radially symmetric heightfields. The symmetry allows to store only a 1-dimensional array of height samples. Non-symmetric heightfields would only make sense, once photon distributions are also recorded in two dimensions. Additionally, each height value is associated with a spacing, to enable irregular grids. The surface between height samples is defined by linear interpolation. In 3-dimensional space, a surface is therefore a sequence of connected concentric capped cones of finite height.

@@

Work distribution is possible using different strategies. One implemented approach assumes equal processing power among MPI nodes and divides the total target number of photons up-front by the number of nodes, before each node starts its part of the simulation (note that we use the term photon, when we actually refer to photon packets, which is explained in \autoref{impl:photon-transport} about photon transport). In practice, this approach did not work well. Specifically, it sometimes caused MPI to abort because of a timeout during Reduce. Therefore a more fine-grained prallelization approach was implemented as follows.

@@

Heights are interpreted from heightfield origin, which is defined by layer thicknesses, into negative z-direction, because in MCML the z-axis points \emph{down} into the material. The last height sample is valid for all radii that are out of bounds.

@@

To calculate reflections and transmissions at heightfield boundaries, a normal is needed. The line-cone intersection method was extended to additionally return the normal at the point of intersection. There are sharp local minima and maxima where the gradient cannot be defined, since the heightfield is interpolated linearly. For these cases the boundary is treated as a xy-plane with normal along z-axis. The direction of the normal is always chosen to point into the half space from where an intersecting ray comes in.

If there was an intersection in the last photon step and the photon is still inside the simulation domain, the corresponding boundary has to be temporarily disabled for intersections during the next step. The reason for this is that the photon might not be moved exactly onto the boundary surface, because of limited floating point accuracy, and it would therefore be possible to detect another erroneous intersection with the same boundary.

@@

where the machinefile mpi.txt specifies the IP address and number of processes per MPI node. The executable is called clustermcml-windows and receives as first argument the MCML kernel file, which will be compiled by the OpenCL implementation. After the kernel file follow the OpenCL compiler arguments, where this example wants to print all warnings as errors. On the tested machines, the current kernel code did not generate warnings. You can add other compiler flags, such as "-D IGNORE\_A" to skip writing to absorption array or "-D IGNORE\_HEIGHTFIELDS" to use implicit flat boundaries, where both options exist to speed up simulation times.

@@

The implementation was currently only built and run on Microsoft Windows, but all used libraries are also available on other platforms. There is no build system other than make (on Windows: nmake) in place, because the project is relatively small and simple. The Makefile can be altered with analog compiler commands and flags, as well as runtime and library paths.

@@

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/verification1.pdf}
	\caption{Comparison of the diffuse reflectance profiles obtained with the implementation presented in this report, clustermcml, and the previous implementation CUDAMCML as reference.}
	\label{verification1}
\end{figure}

@@

\autoref{verification1} compares the diffuse reflectance profiles of clustermcml and CUDAMCML using 100 million photons and one layer of thickness $dl=0.1cm$. The absorption coefficient is $\mu_a=0.1$, the scattering coefficient $\mu_s=1$ and scattering anisotropy $g=0.5$. The refractive indices are $n=1.34$ inside the layer and $1$ outside. The detection window has a size of $1cm$ and a resolution of $0.005cm$, i.e. using $200$ bins. \autoref{verification2} compares transmission profiles of the same simulation.

@@

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/verification2.pdf}
	\caption{Corresponding transmission profiles for~\autoref{verification1}.}
	\label{verification2}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{img/verification3.pdf}
\caption{Diffuse reflectance profiles for a thicker layer than the one in~\autoref{verification1}.}
\label{verification3}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{img/verification4.pdf}
\caption{Corresponding transmission profiles for~\autoref{verification3}.}
\label{verification4}
\end{figure}

\autoref{verification3} and~\autoref{verification4} compare profiles for a thicker layer of $dl=1cm$. The main curve characteristic here is smoothness. Through the low amount of noise, the curves of both implementations almost exactly match each other.

@@

Verification was only possible with flat boundary surfaces. However, to provide validity to some degree with respect to structured boundaries as described in~\autoref{impl:boundary_specification} and~\autoref{impl:intersection}, the same intersection algorithm was used, while input data consisted of degenerated "flat" cones.

@@

The parameters can be read from the legend, which uses the same naming convention as in the previous section. Boundary surfaces are shown as dotted lines in the same color as the corresponding radiance profile. The following shapes were used.

@@

The shapes are the same at the top and bottom boundary of the layer.

@@

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/result1.pdf}
	\caption{Comparison of diffuse reflectance profiles of 7 different surfaces.}
	\label{result1}
\end{figure}

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/result2.pdf}
	\caption{Comparison of diffuse reflectance profiles of flat surface with 4 variations of curved surfaces. Solid lines belong to the left and dotted lines to the right axis.}
	\label{result2}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{img/result3.pdf}
\caption{Comparison of diffuse reflectance profiles of flat surface with 2 variation of noise surfaces. Solid lines belong to the left and dotted lines to the right axis.}
\label{result3}
\end{figure}

@@

\FloatBarrier
\bibliography{bibliography}
\end{document}

>>>>

\documentclass[]{article}
\usepackage[english]{babel} % advanced multi-language features
\usepackage{graphicx} % figures
\usepackage{hyperref} % references and URLs are clickable
\usepackage{cleveref}
\usepackage{listings} % simple code listings (monospaced font, line oriented, no syntax highlighting)
\usepackage[section]{placeins} % restricting floating elements like figures with FloatBarriers
\bibliographystyle{IEEEtranS}

@@

\begin{abstract}
	This project report describes an extended implementation of MCML. It supports structured boundaries, whereas original MCML only supported planar layers. This allows to use the method for simulation of light in more realistic materials. The report presents and compares radiance profiles of different boundary shapes. The focus lies on the properties reflectance and transmittance, that are important for visual appearance.
\end{abstract}

@@

Light transport simulation has uses in many fields nowadays. In astrophysics, it can reveal the material constitution of stars. In medicine, it supports non-invasive diagnostics by providing information for the choice of parameters before operating radiative devices such as lasers or X-rays. Light transport also became important in computer graphics, when synthesizing images of complex materials, where subsurface scattering determines appearance.

Multi-layered Monte Carlo photon transport\cite{wang1992monte} (MCML) is a system for accurate simulations of light transport in materials with complex geometry. This report presents a MCML implementation, that allows even more complex geometry, by specifying the boundaries between layers as heightfields. The purpose is to provide a ground truth of surface appearance for physically based rendering. Monte Carlo methods are accepted to be leading in accuracy, but low-noise results have to be bought with long computation times. The implementation therefore employs massive parallelization.

@@

Monte Carlo simulation of light is very accurate, because it simulates single photon packets and records the distribution of a large number of photons. This implies, that it is a statistical method, using a pseudo random number generator to compute, what can be thought of random walks of photons, visualized in \Cref{trajectory}.

@@

The material geometry in MCML is a sequence of layers as shown in \Cref{layer-model}. Photons enter through the coordinate origin and move in 3 dimensions.

@@

MCML is based on the light propagation model by Prahl et al.~\cite{prahl89}. You can get an overview of the algorithm by looking at the flowchart in \Cref{flowchart}.

@@

Specifics with regard to multiple layers, variable boundary shapes and general implementation are described in \Cref{impl}.

@@

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/radial_heightfield.pdf}
	\caption{A layer boundary surface in the new MCML implementation is represented by a radial heightfield. The samples shown on the left are interpreted in 3-space as shown on the right. In this example, $10$ samples, uniform spacing $0.1cm$ and a total height of $1cm$ are used. Outside the sampled range, boundaries are flat.}
	\label{heightfield}
\end{figure}

The other contribution is support for non-planar layers. This is done by representing boundaries as radially symmetric heightfields. The symmetry allows to store only a 1-dimensional array of height samples. Non-symmetric heightfields would only make sense, once photon distributions are also recorded in two dimensions. To enable irregular grids, eEach height value is associated with a spacing. The surface between height samples is defined by linear interpolation. Thus, in 3-dimensional space, a surface is a sequence of connected concentric capped cones of finite height. ~\Cref{heightfield} shows an example of a heightfield.

@@

Work distribution is possible using different strategies. One implemented approach assumes equal processing power among MPI nodes and divides the total target number of photons up-front by the number of nodes, before each node starts its part of the simulation (note that we use the term photon, when we actually refer to photon packets, which is explained in \Cref{impl:photon-transport} about photon transport). In practice, this approach did not work well. Specifically, it sometimes caused MPI to abort because of a timeout during Reduce. Therefore a more fine-grained prallelization approach was implemented as follows.

@@

Heights are interpreted from heightfield origin, which is defined by layer thicknesses, into negative z-direction (note that in MCML the z-axis points \emph{down} into the material). The last height sample is taken for all radii that are out of bounds.

@@

To calculate reflections and transmissions at heightfield boundaries, a normal is needed. The line-cone intersection method was extended to additionally return the normal at the point of intersection. There are sharp local minima and maxima where the gradient cannot be defined, since the heightfield is interpolated linearly. For these cases the boundary is treated as a xy-plane with normal along z-axis. The direction of the normal is always chosen to point into the half-space from where an intersecting ray comes in.

If there was an intersection in the last photon step and the photon is still inside the simulation domain, the corresponding boundary has to be temporarily disabled for intersections during the next step. The reason for this is that the photon is in most cases not be moved exactly onto the boundary surface, because of limited floating point accuracy, and it would therefore be possible to detect another erroneous intersection with the same boundary.

@@

where the machinefile mpi.txt specifies the IP address and number of processes per MPI node (details can be found in MPI documentation). The executable is called \emph{clustermcml-windows} and receives as first argument the MCML kernel file, which will be compiled by the OpenCL implementation. After the kernel file follow the OpenCL compiler arguments, where this example wants to print all warnings as errors. On the tested machines, the current kernel code did not generate warnings. You can add other compiler flags, such as "-D IGNORE\_A" to skip writing to absorption array or "-D IGNORE\_HEIGHTFIELDS" to use implicit flat boundaries, where both options exist to speed up simulation times.

@@

The implementation was currently only built and run on Microsoft Windows, but all used libraries are also available on other platforms. There is no build system other than make (on Windows: nmake) in place, because the project is relatively small and simple. The Makefile is well structured and can be altered with analog compiler commands and flags, as well as runtime and library paths easily.

@@

@@

\Cref{verification1} compares the radiance profiles of clustermcml and CUDAMCML using 100 million photons and one layer of thickness $dl=0.1cm$. The absorption coefficient is $\mu_a=0.1$, the scattering coefficient $\mu_s=1$ and scattering anisotropy $g=0.5$. The refractive indices are $n=1.34$ inside the layer and $1$ outside. The detection window has a size of $1cm$ and a resolution of $0.005cm$, i.e. using $200$ bins.

@@

\Cref{verification3} compares profiles for a thicker layer of $dl=1cm$. The main curve characteristic here is smoothness. Through the low amount of noise, the curves of both implementations almost exactly match each other.

A simulation with two layers was performed with resulting plots shown in~\Cref{verification5}. Each layer has thickness $dl=1.5cm$, absorption coefficient $\mu_a=0.1$ and scattering coefficient $\mu_s=10$. Refractive indices of $n_1=1.33$ and $n_2=1.2$ and scattering anisotropies of $g_1=0$ and $g_2=0.9$ were used for the first and second layer respectively.

@@

Verification was only possible with flat boundary surfaces. However, to provide validity to some degree with respect to structured boundaries as described in~\Cref{impl:boundary_specification} and~\Cref{impl:intersection}, the same intersection algorithm was used, while input data consisted of degenerated "flat" cones.

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/verification1.pdf}
	\includegraphics[width=\linewidth]{img/verification2.pdf}
	\caption{Comparison of the diffuse reflectance and total transmittance profiles obtained with the implementation presented in this report, clustermcml, and the previous implementation CUDAMCML as reference.}
	\label{verification1}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{img/verification3.pdf}
\includegraphics[width=\linewidth]{img/verification4.pdf}
\caption{Verifying reflectance and transmittance profiles for a thicker layer than the one in~\Cref{verification1} against CUDAMCML.}
\label{verification3}
\end{figure}

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/verification5.pdf}
	\includegraphics[width=\linewidth]{img/verification6.pdf}
	\caption{Verifying reflectance and transmittance profiles for two layers against CUDAMCML.}
	\label{verification5}
\end{figure}

@@

For the simulations, a single layer was used. For demonstration purposes, we chose the following shapes, each used at the top and bottom boundary:

@@

@@

Boundary surfaces are shown as dotted lines in the same color as the corresponding radiance profile. Simulation parameters can be read from the legend, which uses the same naming convention as in the previous section.

\Cref{results:fig:hill} compares the diffuse reflectance and total transmittance profiles of flat with hill-shaped boundaries. \Cref{results:fig:crater} and \Cref{results:fig:noise} are analogous plots with crater-shaped and noise-shaped boundaries, respectively.

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/hill_R.pdf}
	\includegraphics[width=\linewidth]{img/hill_T.pdf}
	\caption{Comparison of flat with hill-shaped boundaries. Reflectance (top) and transmittance (bottom) are shown as solid lines and use the left y-axis. Heightfields are shown as dotted lines and use the right y-axis.}
	\label{results:fig:hill}
\end{figure}

\begin{figure}[ht!]
	\includegraphics[width=\linewidth]{img/crater_R.pdf}
	\includegraphics[width=\linewidth]{img/crater_T.pdf}
	\caption{Comparison of flat with crater-shaped boundaries.}
	\label{results:fig:crater}
\end{figure}

\begin{figure}[ht!]
\includegraphics[width=\linewidth]{img/noise_R.pdf}
\includegraphics[width=\linewidth]{img/noise_T.pdf}
\caption{Comparison of flat with noise-shaped boundaries.}
\label{results:fig:noise}
\end{figure}

@@

The new implementation of MCML enables us to obtain radiance profiles for a wide range of surface shapes. It can replace measurements as a cheaper way of obtaining ground truths about light distribution.

Radiance profiles reveal the effect of varying surface shapes, and therefore allow to model surface appearance with approximate analytic curves or lookup tables. The intended main benefit of the new implementation is the use of these profiles by fast rendering techniques for subsurface scattering.

Real-world surfaces are often more complex than radially symmetric heightfields. Possible future work is therefore extending the boundary geometry to arbitrary meshes. In this case one should also record radiance in two spatial dimensions.

The decision if a photon passes a boundary or not is currently purely based on Fresnel reflectance. Another subject for future work would be to support more complex BRDFs.

\bibliography{bibliography}
\end{document}
